Mixture Models for Single--Cell Assays with Application to Vaccine Studies

### 2012-09-14 MIMOSA.tex
Final Revisions From Raphael

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
\end{keywords}
\maketitle

@@

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

@@

Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}. 
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points: day 0 and day 182. 
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

@@

Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e. antigen or condition) at a time vs. unstimulated. The issue of multiple antigens is handled through multiple testing correction.

@@

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable~\citep{Bernardo:1996tl} Beta priors on the unknown proportions, as follows: 
 \begin{align*}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u),
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects. 
Note that the parameters $\alpha_u \text{ and } \beta_u$ are explicitly shared across the two models, whereas $\alpha_s \text{ and } \beta_s$ are only present in the alternative model. 
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders. 
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p_{ui}$ and $p_{si}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$. 
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}. 
As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha_u, \beta_u$, $\alpha_s,\text{ and }\beta_s$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1. 
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder by the model (or a competing method).

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.ps}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.ps}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
  \node [below=of B] (E) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.ps}};
    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {E} ;
    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2. ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

@@

\begin{figure}
\centering
\begin{tikzpicture} [auto,node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.ps}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.ps}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\node [below=of A] (C) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.ps}};
\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The posterior differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Clustering of the signed q-values from Fisher's exact test. Genes selected from Fisher's exact test at the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}

@@

For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \begin{tikzpicture} [auto, node distance=0cm]
   \node at (0,0) (A) {
\begin{tikzpicture}
    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.ps}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {A} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [right=of A] (B) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.ps}};
    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
    \end{tikzpicture}
    };
        \node [below=of A] (C) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.ps}};
    \begin{scope}[x={(c.south east)},y={(c.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.ps}};
    \begin{scope}[x={(d.south east)},y={(d.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
    };

@@

\subsection{Prior}
As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, such that
\begin{align*}
(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\\\nonumber
(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s),\end{align*}
where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$. As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation. When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.

@@

\begin{figure}
\centering
\begin{tikzpicture} [auto, node distance=0 cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.ps}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B){
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.ps}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.02,1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
%\node at (0,0) [font=\small\sffamily] {C} ;
%\node at (6.25,0) [font=\small\sffamily] {D} ;
\end{tikzpicture}
\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions. No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples. Sample names highlighted in red identify those where MIMOSA did not detect a difference. This figure appears in color in the electronic version of this article.}
\label{fig:polyfunctionality}
\end{figure}

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.ps}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.ps}};
   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
       \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
   \caption{Multivariate simulations from a two-sided model. Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500). Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test. A) Average ROC curves for the competing methods over 10 simulations. B) Average observed and nominal false discovery rate for each method over 10 simulations. This figure appears in color in the electronic version of this article.}
   \label{fig:mvsimulations}
\end{figure}

>>>>

\begin{keywords}
Expectations-Maximization, Markov Chain Monte Carlo, Marginal
Likelihood, Bayesian Modeling, Hierarchical Modeling, Immunology, Flow
Cytometry, Single-Cell Gene Expression
\end{keywords}
\maketitle

@@

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

@@

Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

@@

Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e. antigen or condition) at a time vs. unstimulated.

@@

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
 \begin{align*}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
(p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u,
\beta_u) \quad\mathrm{and}\quad (p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s),
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects.
Note that the parameters $\alpha_u \text{ and } \beta_u$ are explicitly shared across the two models, whereas $\alpha_s \text{ and } \beta_s$ are only present in the alternative model.
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p_{ui}$ and $p_{si}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha_u, \beta_u$, $\alpha_s,\text{ and }\beta_s$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
  \node [below=of B] (E) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {E} ;
    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2. ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

@@

\begin{figure}
\centering
\begin{tikzpicture} [auto,node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\node [below=of A] (C) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The posterior differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Clustering of the signed q-values from Fisher's exact test. Genes selected from Fisher's exact test at the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}

@@

For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \begin{tikzpicture} [auto, node distance=0cm]
   \node at (0,0) (A) {
\begin{tikzpicture}
    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {A} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [right=of A] (B) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
    \end{tikzpicture}
    };
        \node [below=of A] (C) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
    \begin{scope}[x={(c.south east)},y={(c.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
    \begin{scope}[x={(d.south east)},y={(d.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
    };

@@

\subsection{Prior}
As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, such that
\begin{align*}
(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\\\nonumber
 (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s) \quad \text{and}\quad (\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\end{align*}
where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$. As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation. When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.

@@

\begin{figure}
\centering
\begin{tikzpicture} [auto, node distance=0 cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B){
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.02,1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
%\node at (0,0) [font=\small\sffamily] {C} ;
%\node at (6.25,0) [font=\small\sffamily] {D} ;
\end{tikzpicture}
\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions. No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples. Sample names highlighted in red identify those where MIMOSA did not detect a difference. This figure appears in color in the electronic version of this article.}
\label{fig:polyfunctionality}
\end{figure}

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
       \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
   \caption{Multivariate simulations from a two-sided model. Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500). Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test. A) Average ROC curves for the competing methods over 10 simulations. B) Average observed and nominal false discovery rate for each method over 10 simulations. This figure appears in color in the electronic version of this article.}
   \label{fig:mvsimulations}
\end{figure}

\section{Discussion}
\label{s:discussion}
Experimentalists already have access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few. As single-cell assays become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important. Current approaches for single-cell assays are for the most part simplistic such as the t-test, $\chi^2$ test, and Fisher's exact test, and resulting inference can be quite sub-optimal, especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy,Gottardo:2006}. In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.
